---
output:
  pdf_document: default
  html_document: default
---

# Interactions, Conditioning

```{r 7.1}
library(rethinking)
data(rugged)
d <- rugged

## make log version of outcome
d$log_gdp <- log(d$rgdppc_2000)

## extract countries with GDP data
dd <- d[complete.cases(d$rgdppc_2000),]

## split countries into Africa and not-Africa
d.A1 <- dd[dd$cont_africa==1,] # Africa
d.A0 <- dd[dd$cont_africa==0,] # not Africa
```

```{r 7.2}
# African nations
m7.1 <- map(
    alist(
        log_gdp ~ dnorm(mu, sigma),
        mu <- a + bR*rugged,
        a ~ dnorm(8, 100),
        bR ~ dnorm(0, 1),
        sigma ~ dunif(0, 10)
    ),
    data=d.A1)

## non-African nations
m7.2 <- map(
    alist(
        log_gdp ~ dnorm(mu, sigma),
        mu <- a + bR*rugged,
        a ~ dnorm(8, 100),
        bR ~ dnorm(0, 1),
        sigma ~ dunif(0, 10)
    ),
    data=d.A0)
```

```{r fig-7.2}
par(mfrow=c(1,2))

rugged.seq <- seq(from=-0.5, to=6.5, length.out=30)
pred_dat <- list(rugged=rugged.seq)
mu <- link(m7.1, data=pred_dat)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI, prob=0.89)

plot(log(rgdppc_2000) ~ rugged, d.A1, col=col.alpha(rangi2,1))
mtext("Africa", 3)
lines(rugged.seq, mu.mean)
shade(mu.PI, rugged.seq)

mu <- link(m7.2, data=pred_dat)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI, prob=0.89)

plot(log(rgdppc_2000) ~ rugged, d.A0, col="black")
mtext("Not Africa", 3)
lines(rugged.seq, mu.mean)
shade(mu.PI, rugged.seq)
```

```{r 7.3}
m7.3 <- map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR*rugged,
    a ~ dnorm(8, 100),
    bR ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data=dd)
```

```{r 7.4}
m7.4 <- map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + bR*rugged + bA*cont_africa,
    a ~ dnorm(8, 100),
    bR ~ dnorm(0, 1),
    bA ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data=dd)
```

```{r 7.5}
compare(m7.3, m7.4)
```

`m7.4` gets all the model weight and the difference in WAIC is large (and beyond the bounds of the SE -- `63 \pm 15`). So there it is likely that the effect of being an African nation is important.

```{r 7.6}
rugged.seq <- seq(from=-1, to=8, by=0.25)

## compute mu over samples, fixing cont_africa=0
mu.NotAfrica <- link(m7.4, data=data.frame(cont_africa=0, rugged=rugged.seq))

## compute mu over sample, fixing cont_africa=1
mu.Africa <- link(m7.4, data=data.frame(cont_africa=1, rugged=rugged.seq))

## summarize to means and intervals
mu.NotAfrica.mean <- apply(mu.NotAfrica, 2, mean)
mu.NotAfrica.PI <- apply(mu.NotAfrica, 2, PI, prob=0.97)
mu.Africa.mean <- apply(mu.Africa, 2, mean)
mu.Africa.PI <- apply(mu.Africa, 2, PI, prob=0.97)
```

```{r fig-7.3}
par(mfrow=c(1,1))
plot(log_gdp ~ rugged, data=dd, col=NULL,
     xlab="Terrain Ruggedness Index",
     ylab="log GDP year 2000")

points(log_gdp ~ rugged, data=subset(dd, cont_africa == 0), col=col.alpha("black", 0.6))
points(log_gdp ~ rugged, data=subset(dd, cont_africa == 1), col=col.alpha(rangi2, 0.6))

lines(rugged.seq, mu.NotAfrica.mean)
shade(mu.NotAfrica.PI, rugged.seq)
text(x=5.5, y=9.4, labels="not Africa")

lines(rugged.seq, mu.Africa.mean, col=rangi2)
shade(mu.Africa.PI, rugged.seq, col=col.alpha(rangi2, 0.3))
text(x=4.2, y=6.5, labels="Africa", col=rangi2)
```

The dummy variable does not allow the slope to change conditional on continent. The model `m7.4` seems to only reflect that African nations have lower average GDP.

Now we make the effect of ruggedness conditional on africa also:

```{r 7.7}
m7.5 <- map(
  alist(
    log_gdp ~ dnorm(mu, sigma),
    mu <- a + gamma*rugged + bA*cont_africa,
    gamma <- bR + bAR*cont_africa,
    a ~ dnorm(8, 100),
    bA ~ dnorm(0, 1),
    bR ~ dnorm(0, 1),
    bAR ~ dnorm(0, 1),
    sigma ~ dunif(0, 10)
  ),
  data=dd)
```

```{r 7.8}
compare(m7.3, m7.4, m7.5)
```

```{r 7.10}
rugged.seq <- seq(from=-1, to=8, by=0.25)

mu.Africa <- link(m7.5, data=data.frame(cont_africa=1, rugged=rugged.seq))
mu.Africa.mean <- apply(mu.Africa, 2, mean)
mu.Africa.PI <- apply(mu.Africa, 2, PI, prob=0.97)

mu.NotAfrica <- link(m7.5, data=data.frame(cont_africa=0, rugged=rugged.seq))
mu.NotAfrica.mean <- apply(mu.NotAfrica, 2, mean)
mu.NotAfrica.PI <- apply(mu.NotAfrica, 2, PI, prob=0.97)
```

```{r 7.11}
par(mfrow=c(1,2))

## plot African nations with regression
d.A1 <- dd[dd$cont_africa==1,]
plot(log(rgdppc_2000) ~ rugged,
     data=d.A1,
     col=rangi2,
     ylab="log GDP year 2000",
     xlab="Terrain Ruggedness Index")
mtext("African nations", 3)
lines(rugged.seq, mu.Africa.mean, col=rangi2)
shade(mu.Africa.PI, rugged.seq, col=col.alpha(rangi2, 0.3))

## plot non-African nations with regresion
d.A0 <- dd[dd$cont_africa==0,]
plot(log(rgdppc_2000) ~ rugged,
     data=d.A0,
     col="black",
     ylab="log GDP year 2000",
     xlab="Terrain Ruggedness Index")
mtext("Non-African nations", 3)
lines(rugged.seq, mu.NotAfrica.mean)
shade(mu.NotAfrica.PI, rugged.seq)
```

```{r 7.12}
precis(m7.5)
```
```{r 7.13}
post <- extract.samples(m7.5)
gamma.Africa <- post$bR + post$bAR*1
gamma.notAfrica <- post$bR + post$bAR*0
```

```{r 7.14}
mean(gamma.Africa)
mean(gamma.notAfrica)
```

```{r 7.15}
par(mfrow=c(1,1))
dens(gamma.Africa, xlim=c(-0.5, 0.5), ylim=c(0,5.5),
     xlab="gamma", col=rangi2)
text(x=0.28, y=3.75, "Africa", col=rangi2)

dens(gamma.notAfrica, add=TRUE)
text(x=-0.07, y=5.1, "not Africa")

mtext("Slope", 3)
```

```{r 7.16}
diff <- gamma.Africa - gamma.notAfrica
sum(diff < 0) / length(diff)
```

```{r 7.17}
## get minimum and maximum rugged values
q.rugged <- range(dd$rugged)

## compute lines and confidence intervals
mu.ruggedlo <- link(m7.5,
                    data=data.frame(
                      rugged=q.rugged[1],
                      cont_africa=0:1))
mu.ruggedlo.mean <- apply(mu.ruggedlo, 2, mean)
mu.ruggedlo.PI <- apply(mu.ruggedlo, 2, PI)

mu.ruggedhi <- link(m7.5,
                    data=data.frame(
                      rugged=q.rugged[2],
                      cont_africa=0:1))
mu.ruggedhi.mean <- apply(mu.ruggedhi, 2, mean)
mu.ruggedhi.PI <- apply(mu.ruggedhi, 2, PI)

## plot it all, splitting points at median
med.r <- median(dd$rugged)
ox <- ifelse(dd$rugged > med.r, 0.05, -0.05)
plot(dd$cont_africa + ox, log(dd$rgdppc_2000),
     col=ifelse(dd$rugged > med.r, rangi2, "black"),
     xlim=c(-0.25,1.25), xaxt="n", ylab="log GDP year 2000",
     xlab="Continent")
axis(1, at=c(0,1), labels=c("other", "Africa"))
lines(0:1, mu.ruggedlo.mean, lty=2)
shade(mu.ruggedlo.PI, 0:1)
lines(0:1, mu.ruggedhi.mean, col=rangi2)
shade(mu.ruggedhi.PI, 0:1, col=col.alpha(rangi2,0.25))
```

## 7.3 Continuous interactions

```{r 7.18}
library(rethinking)
data(tulips)
d <- tulips
str(d)
```
