```{r 2.1}
ways <- c(0, 3, 8, 9, 0)
ways / sum(ways)
```

```{r 2.2}
dbinom(6, size=9, prob=0.5)
```

```{r 2.3}
## define grid
p_grid <- seq(from=0, to=1, length.out=20)

## define prior
prior <- rep(1, 20)

## compute likelihood at each value in grid
likelihood <- dbinom(6, size=9, prob=p_grid)

## compute product of likelihood and prior
unstd.posterior <- likelihood * prior

## standardize the posterior, so it sums to 1
posterior <- unstd.posterior / sum(unstd.posterior)
```

```{r 2.4}
plot(p_grid, posterior, type="b",
     xlab="probability of water", ylab="posterior probability")
```

```{r 2.6}
library(rethinking)
globe.qa <- map(
  alist(
    w ~ dbinom(9,p),                    # binomial likelihood
    p ~ dunif(0,1)                      # uniform prior
  ),
  data=list(w=6))

precis(globe.qa)
```

```{r 2.7}
# analytical calculation
w <- 6
n <- 9
curve(dbeta(x, w+1, n-w+1), from=0, to=1)
# quadratic approximation
curve(dnorm(x, 0.67, 0.16), lty=2, add=TRUE)
```

## 2.6 Practice

### Easy

  - 2E1. Pr(rain|Monday), Pr(rain,Monday) / Pr(Monday), (the
    probability of rain on Monday)
  - 2E2. (3) The probability that it is Monday, given that it is
    raining (Pr(Monday|rain))
  - 2E3. (1) Pr(Monday|rain), (4) Pr(rain|Monday) Pr(Monday) /
    Pr(rain) (the probability that it is Monday, given that it is
    raining)
  - 2E4. "The probability of water is 0.7" describes a belief (a
    willingness to bet, for instance) about the chances/likelihood
    that a single sample of a point on the globe will result in water.

### Medium

#### 2M1

Compute/plot grid approximation of posterior distribution for each set of observations

```{r ex-2M1}
grid_approx <- function(obs, prior) {
  obs_list <- strsplit(obs, " ")[[1]]
  nwater <- sum(obs_list == 'W')
  n <- length(obs_list)

  p_grid <- seq(from=0, to=1, length.out=20)
  likelihood <- dbinom(nwater, size=n, prob=p_grid)
  unstd.posterior <- likelihood * sapply(p_grid, prior)
  posterior <- unstd.posterior / sum(unstd.posterior)

  plot(p_grid, posterior, type='b',
       xlab="probability of water", ylab="posterior probability")
  title(obs)
}

samples <- c(
  "W W W",
  "W W W L",
  "L W W L W W W"
)
par(mfrow=c(1, 3))

prior <- function(p) { 1 }
sapply(samples, function(obs) {
  grid_approx(obs, prior)
  })
```

### 2M2

```{r ex-2M2}
prior <- function(p) {
  if (p < 0.5)  0
  else          1
}
sapply(samples, function(obs) {
  grid_approx(obs, prior)
  })
```

### 2M3

```{r ex-2M3}
p_land_earth <- 0.3
p_land_mars <- 1

p_earth <- 0.5
p_mars <- 0.5
p_land <- sum(c(p_land_earth, p_land_mars) * c(p_earth, p_mars))

# pr(earth|land) <- (pr(land|earth) pr(earth)) / pr(land)
p_earth_land <- (p_land_earth * p_earth) / p_land
p_earth_land
```

### 2M4

```{r ex-2M4}
p_card <- 1/3
p_black <- sum(c(1, 0.5, 0) * p_card)
p_black_c2 <- 1
# pr(two-black|black) <- (pr(black|two-black) * pr(two-black)) / pr(black)
p_c2_black <- (p_black_c2 * p_card) / p_black
p_c2_black
```

### 2M5

```{r ex-2M5}
p_card <- 1/4
p_black <- sum(c(1, 1/2, 0, 1) * p_card)
p_c2_black <- (p_black_c2 * p_card) / p_black
p_c2_black
```

### 2M6

```{r ex-2M6}
p_card <- c(1/6, 2/6, 3/6)
sides_black <- c(2, 1, 0)
names(p_card) <- c('c2', 'c1', 'c0')
p_black <- sum(sides_black/2 * p_card)
p_c2_black <- (p_black_c2 * p_card['c2']) / p_black
p_c2_black
```

### Hard

#### 2H1

```{r ex-2H1}
# P(Twin | SpeciesA) = P(T|A) = 0.1
# P(Twin | SpeciesB) = P(T|B) = 0.2
# P(A) = P(B) = 0.5
p_t_a <- 0.1
p_t_b <- 0.2
p_a <- 0.5
p_b <- 0.5

# total probability of a twin
# P(T) = P(T|A) * P(A) + P(T|B) * P(B)
p_twin <- p_t_a * p_a + p_t_b * p_b

# conditional probability of species, given observing a twin
p_a_t <- (p_t_a * p_a) / p_twin
p_b_t <- (p_t_b * p_b) / p_twin

# probability of twin, given a twin
# P(T|T) = P(T|A)*P(T,A) + P(T|B)*P(T,B)
p_t_t <- p_t_a * p_a_t + p_t_b * p_b_t
p_t_t
```

The probability is somewhere between the two priors (0.1 and 0.2), but
slightly favoring species B.

#### 2H2

```{r ex-2H2}
# P(A|T) = P(T|A)*P(A) / P(T), calculated above
p_a_t
```

#### 2H3

```{r ex-2H3}
p_s_a <- 1 - p_t_a
p_s_b <- 1 - p_t_b
# P(Single) = P(S) = P(S|A)*P(A) + P(S|B)*P(B)
# conditioned on having twin first,
#   P(S|T) = P(S|A)*P(A,T) + P(S|B)*P(B,T)
p_s_t <- p_s_a * p_a_t + p_s_b * p_b_t

# P(A|TS) = P(S|A)*P(A|T) / P(S|T)
p_a_ts = (p_s_a * p_a_t) / p_s_t
p_a_ts
```

### 2H4

```{r ex-2H4}
# no prior information on births
# P(TestA|A) = 0.8
# P(TestB|B) = 0.65
p_tA_a <- 0.8
p_tB_b <- 0.65
p_tA_b <- 1 - p_tB_b

# P(TestA) = P(TestA|A)*P(A) + P(TestA|B)*P(B)
p_tA <- p_tA_a * 0.5 + (1 - p_tB_b) * 0.5

# P(A|TestA) = P(TestA|A)*P(A) / P(TestA)
p_a_tA <- p_tA_a * 0.5 / p_tA
p_a_tA

# with priors on births
# P(TestA|TS) = P(TestA|A)*P(A|TS) + P(TestA|B)*P(B|TS)
p_b_ts <- (p_s_b * p_b_t) / p_s_t
p_tA_ts <- p_tA_a * p_a_ts + p_tA_b * p_b_ts

# P(A|TestA,TS) = P(TestA|A)*P(A|TS) / P(TestA|TS)
p_a_tAts <- p_tA_a * p_a_ts / p_tA_ts
p_a_tAts
```
