# Statistical Rethinking, Chapter 3

Conditional probability of positive result on test, given rare underlying probability of event

```{r 3.1}
PrPV <- 0.95                            # P(positive|vampire)
PrPM <- 0.01                            # P(positive|mortal)
PrV <- 0.001                            # P(vampire)
PrP <- PrPV*PrV + PrPM*(1-PrV)          # P(positive) = [total prob.]
( PrVP <- PrPV*PrV / PrP )              # P(vampire|positive)
```

## 3.1 Sampling form grid-approximate posterior

```{r 3.2}
# grid approximation
p_grid <- seq(from=0, to=1, length.out=1000)
prior <- rep(1, 1000)
likelihood <- dbinom(6, size=9, prob=p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
```

sampling from the posterior

```{r 3.3}
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
```

```{r 3.4}
par(mfcol=c(1,2))
plot(samples, col=alpha("royalblue2", 0.3), pch=16,
     xlab="sample number", ylab="proportion water (p)")
```

```{r 3.5}
library(rethinking)
dens(samples, lwd=1.5, col="royalblue2", xlab="proportion water (p)")
```

## 3.2 Sampling to summarize

```{r 3.6}
# add up posterior from grid approximation
sum(posterior[p_grid < 0.5])
```

```{r 3.7}
# defined boundary interval, upper bound
sum(samples < 0.5) / 1e4
```

```{r 3.8}
# defined boundary interval, region
sum(samples > 0.5 & samples < 0.75) / 1e4
```

```{r 3.9}
# defined mass interval, upper bound
quantile(samples, 0.8)
```

```{r 3.10}
# defined mass interval, region
quantile(samples, c(0.1, 0.9))
```

### Percentile Intervals can be misleading

Percentile intervals assign equal probability mass to each tail. For example, a 95% percentile interval (often called a __confidence interval_) excludes 2.5% probability mass on each side.

This will work poorly if the distribution is highly skewed. For example,

```{r 3.11}
p_grid <- seq(from=0, to=1, length.out=1000)
prior <- rep(1, 1000)
likelihood <- dbinom(3, size=3, prob=p_grid) # 3 successes in 3 trials
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
samples <- sample(p_grid, size=1e4, replace=TRUE, prob=posterior)
```

The 50% percentile interval (PI) provides the bounds on a central 50% probability. But the interval excludes the most probable values in this skewed distribution

```{r 3.12}
PI(samples, prob=0.5)
```

By contrast, the 50% highest posterior density interval (HPDI) is the narrowest interval that contains 50% mass. It will choose the 'densest' possible region.

```{r 3.13}
HPDI(samples, prob=0.5)
```

```{r fig3.3}
# recreate figure 3.3
par(mfrow=c(1,2))

plot(p_grid, posterior, type='l',
     xlab="proportion water (p)", xaxp=c(0, 1, 4),
     ylab="density")
pi <- PI(samples, prob=0.5)
coord.select <- p_grid >= pi[1] & p_grid <= pi[2]
coord.x <- c(pi[1], p_grid[coord.select], pi[2])
coord.y <- c(0, posterior[coord.select], 0)
polygon(coord.x, coord.y, col="royalblue2", border=NA)
title('50% Percentile Interval')

plot(p_grid, posterior, type='l',
     xlab="proportion water (p)", xaxp=c(0, 1, 4),
     ylab="density")
hpdi <- HPDI(samples, prob=0.5)
coord.select <- p_grid >= hpdi[1] & p_grid <= hpdi[2]
coord.x <- c(hpdi[1], p_grid[coord.select], hpdi[2])
coord.y <- c(0, posterior[coord.select], 0)
polygon(coord.x, coord.y, col="royalblue2", border=NA)
title('50% HPDI')
```

### 3 point estimates: mean, median, mode

_maximum a posteriori (MAP)_ estimate, by grid:

```{r 3.14}
# MAP/mode
p_grid[which.max(posterior)]
```

_maximum a posteriori (MAP)_ estimate, by sampling:

```{r 3.15}
# from rethinking package
chainmode(samples, adj=0.01)
```

mean or median:

```{r 3.16}
mean(samples)
median(samples)
```

## Loss functions

Let decision $d = 0.5$ be the guess that the true value of parameter $p$ is $0.5$. Then choose the loss to be $d - p$ (the absolute loss), thus _weighted average loss_ is:

```{r 3.17}
sum(posterior*abs(0.5-p_grid))
```

where the weights are the posterior probabilities for each loss given each parameter $p$ in vector `p_grid`.


Getting losses for all decisions:

```{r 3.18}
## all decisions
loss <- sapply(p_grid, function(d) sum(posterior*abs(d-p_grid)))
```

Find the loss-minimizing decision (which will be the median):

```{r 3.19}
p_grid[which.min(loss)]
# == median(samples)
```

Choice of loss function, most common:

  * absolute loss ($d - p$) --> median
  * quadratic loss ($(d - p)^2$) --> mean


```{r plot-loss}
par(mfrow=c(1,1))
abs_losses <- loss
plot(p_grid, abs_losses, type='l')  # absolute loss
points(p_grid[which.min(abs_losses)],
       abs_losses[which.min(abs_losses)], col='red')
points(median(samples),
       abs_losses[which.min(abs_losses)], col='royalblue2')

quad_loss <- function(d) sum(posterior * (d - p_grid)^2)
quad_losses <- sapply(p_grid, quad_loss)
lines(p_grid, quad_losses, type='l')
points(p_grid[which.min(quad_losses)],
       quad_losses[which.min(quad_losses)], col='red')
points(mean(samples),
       quad_losses[which.min(quad_losses)], col='royalblue2')

ylim(c(0, 1))
```

## Sampling for simulation

Likelihood for globe tossing model is binomial likelihood

```{r 3.20}
# P(w | n,p) -- probability of water, given n tosses and proportion p = 0.7
dbinom(0:2, size=2, prob=0.7)
```

```{r 3.21}
# a single observation of a 2 toss trial
rbinom(1, size=2, prob=0.7)
```

```{r 3.22}
# 10 observations of a 2 toss trial
rbinom(10, size=2, prob=0.7)
```

```{r 3.23}
# compare simulation frequencies to analytical solution in 3.20
dummy_w <- rbinom(1e5, size=2, prob=0.7)
table(dummy_w)/1e5
```

```{r 3.24}
# increase to 9 tosses per trial
dummy_w <- rbinom(1e5, size=9, prob=0.7)
simplehist(dummy_w, xlab="dummy water count")
```

```{r 3.25}
# simulating predicted observations given known p=0.6
w <- rbinom(1e4, size=9, prob=0.6)
```

```{r 3.26}
# using posterior samples to drive probabilities for simulation
w <- rbinom(1e4, size=9, prob=samples)
```


## Exercises

### Easy

```{r 3.27}
# setup for Easy exercises
p_grid <- seq(from=0, to=1, length.out=1000)
prior <- rep(1, 1000)
likelihood <- dbinom(6, size=9, prob=p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
```

#### 3E1

How much posterior probability lies below p = 0.2?

```{r ex-3E1}
sum(samples < 0.2) / 1e4
```

#### 3E2

How much posterior probability lies above p = 0.8?

```{r ex-3E2}
sum(samples > 0.8) / 1e4
```

#### 3E3

How much posterior probability lies between p = 0.2 and p = 0.8?

```{r ex-3E3}
sum(samples > 0.2 & samples < 0.8) / 1e4
```

#### 3E4

20% of the posterior probability lies below which value of p?

```{r ex-3E4}
quantile(samples, 0.2)
```

#### 3E5

20% of the posterior probability lies above which value of p?

```{r ex-3E5}
quantile(samples, 0.8)
```

#### 3E6

Which values of p contain the narrowest interval equal to 66% of the posterior probability?

```{r ex-3E6}
HPDI(samples, 0.66)
```

#### 3E7

Which values of p contain 66% of the posterior probability, assuming equal posterior probability both below and above the interval?

```{r ex-3E7}
PI(samples, 0.66)
```


### Medium

#### 3M1

Adjust grid approximation code from above, for 8 water in 15 tosses

```{r ex-3M1}
w <- 8
n <- 15
p_grid <- seq(from=0, to=1, length.out=1000)  # grid
prior <- rep(1, 1000)                         # flat prior
likelihood <- dbinom(w, size=n, prob=p_grid) # 8 water in 15 tosses
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
```

#### 3M2

Sample from the posterior, and calculate the 90% HPDI for p

```{r ex-3M2
trials <- 1e4
set.seed(100)
samples <- sample(x=p_grid, prob=posterior, size=trials, replace=TRUE)
HPDI(samples, 0.9)
```

```{r}
plot(p_grid, posterior, type='l',
     xlab="proportion water (p)", xaxp=c(0, 1, 4),
     ylab="density")
hpdi <- HPDI(samples, prob=0.9)
coord.select <- p_grid >= hpdi[1] & p_grid <= hpdi[2]
coord.x <- c(hpdi[1], p_grid[coord.select], hpdi[2])
coord.y <- c(0, posterior[coord.select], 0)
polygon(coord.x, coord.y, col="royalblue2", border=NA)
title('90% HPDI')
```

#### 3M3

Posterior predictive check

```{r ex-3M3}
w <- rbinom(trials, size=n, prob=samples)
mean(w == 8) # probability of 8 water in 15 tosses
```

#### 3M4

Probability of 6 water in 9 tosses, given a prior from previous data: 8 water in 15 tosses.

```{r ex-3M4}
w2 <- rbinom(trials, size=9, prob=samples)
simplehist(w2)
mean(w2 == 6)
```

#### 3M5

```{r ex-3M5}
p_grid <- seq(from=0, to=1, length.out=1000)

prior3 <- rep(1, 1000)
prior3[p_grid < 0.5] <- 0

likelihood3 <- dbinom(8, size=15, prob=p_grid)
posterior3 <- likelihood3 * prior3
posterior3 <- posterior3 / sum(posterior3)
set.seed(100)
samples3 <- sample(p_grid, prob=posterior3, size=1e4, replace=TRUE)

# 3M2'
HPDI(samples3, 0.9)

# 3M3'
w3 <- rbinom(1e4, size=15, prob=samples3)
mean(w3 == 8) # probability of 8 water in 15 tosses

# 3M4'
w4 <- rbinom(1e4, size=9, prob=samples3)
simplehist(w4)
mean(w4 == 6)
```

```{r}
par(mfrow=c(1,2))
simplehist(w3, xlim=c(0, 15))
simplehist(w4, xlim=c(0, 15))
```

### Hard

```{r 3.29}
data(homeworkch3)
```

#### 3H1

```{r ex-3H1}
p_grid <- seq(from=0, to=1, length.out=1000)

prior <- rep(1, 1000)                   # flat prior

nboy <- sum(birth1 == 1) + sum(birth2 == 1)
n <- length(birth1) + length(birth2)
likelihood <- dbinom(nboy, size=n, prob=p_grid)
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)

# two ways to calculate mode
chainmode(samples)
p_grid[which.max(posterior)]
```

#### 3H2

```{r ex-3H2}
set.seed(100)
samples <- sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
HPDI(samples, prob=c(0.5, 0.89, 0.97))
```

#### 3H3

```{r ex-3H3}
boys <- rbinom(1e4, 200, prob=samples)
dens(boys, col='royalblue2', show.HPDI=0.95)
abline(v=111, col='gray')
```

#### 3H4

```{r ex-3H4}
boys2 <- rbinom(1e4, size=100, prob=samples)

dens(boys2, col="royalblue2")
abline(v=sum(birth1 ==  1))
```

#### 3H5

```{r ex-3H5}
BOY <- 1
GIRL <- 0
girls_1st <- sum(birth1 ==  GIRL)
boy_after_girl_ppc <- rbinom(1e4, size=girls_1st, prob=samples)

simplehist(boy_after_girl_ppc, adj=0.1, col='royalblue2')
boy_after_girl_actual <- birth2[birth1 == GIRL]
abline(v=sum(boy_after_girl_actual), col='gray')
```
